---
title: "What is Data Science?"
format: html
toc: TRUE
editor_options: 
  chunk_output_type: inline
---

```{r}
#| include: false
library(rquiz)
```

First, let's get some perspectives on what data science is, how statistics and computer science relate, and important items to consider when trying to understand data. To do so, we'll do some readings.

- <a href="https://r4ds.hadley.nz/intro" target="_blank">Introduction section of *R for Data Science*</a>
- <a href="https://scientistcafe.com/ids/introduction#introduction" target="_blank">Chapters 1 and 2 of *The Practitioner's Guide to Data Science*</a>
- <a
      href="https://www.amstat.org/docs/default-source/amstat-documents/the-role-of-statistics-in-data-science-and-artificial-intelligence.pdf"
      target="_blank">The role of statistics in Data Science and Artificial Intelligence</a>
- <a href="https://en.wikipedia.org/wiki/Data_science" target="_blank">Data Science wikipedia entry</a>

```{r, echo =FALSE, eval = FALSE}
rquiz::multiQuestions(
  x = list(
    q1 = list(
      question = "According to the readings, what are some of the common mistakes a data science project may have?",
      options = c("Solving the wrong problem", 
                  "Using too much data",
                  "Overfitting and obsession with complicated models",
                  "Fail to scale in real-time applications",
                  "Finding an answer that goes against the company's plan",
                  "Too optimistic abou the timeline"
                  ),
      answer = c(1, 3, 4, 6),
      tip = "Check the Practioner's guide to data science!"
    ),
    q2 = list(
      question = "According to R for Data Science, which is the correct order for the steps in the data science process?",
      options = c("Import, Tidy, Understand, Communicate",
                  "Tidy, Import, Understand, Communicate",
                  "Communicate, Tidy, Import, Understand",
                  "Understand, Import, Tidy, Communicate"),
      answer = 1
    )
  ),
  title = "Reading recap",
#  width = "50%",
#  fontSize = 14,
  shuffle = TRUE
)
```

<!-- Note: The questions above are just there to help you check in. These don't count toward your grade! -->

You'll write about these ideas as part of your first homework assignment! For now, onto understanding data workflows and `Git` and `GitHub` basics!

Use the table of contents on the left or the arrows at the bottom of this page to navigate to the next learning material!

