---
title: "Dynamic UIs"
author: "Justin Post"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, "css/ncsu.css", "css/ncsu-fonts.css", "css/mycss.css"]
    nature:
      beforeInit: ["js/ncsu-scale.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
    includes:
      in_header: "partials/header.html"
editor_options: 
  chunk_output_type: console
---


```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
options(dplyr.print_min = 5)
library(knitr)
library(readr)
library(plotly)
library(shiny)
library(DT)
opts_chunk$set(echo = FALSE)
```


layout: false

# Recap

`ui`    

  - Controls layout of app    
  - Basic layout uses a sidebar panel and a main panel
  
      + `bslib::page_sidebar()` or `shiny::sidebarLayout()`
      
  - Use strings, formatted (html style) text, widgets (`*Input` functions), and output from `server` (`*Output` functions)    
  
`server`    

  - Back-end for app      
  - Create outputs that react to inputs (must be a reactive context!)
  
      + `render*` functions, `reactive()`, and `reactiveValues()`
      + `observe()`, `observeEvent()`, and `eventReactive()`


---

#  Dynamic UI  
  
- Often want to update UI based on user input!  
  
- Methods for updating UI  

    + `update*` functions
    + `renderUI()`/`uiOutput()`
    + `conditionalPanel()`


---

#  Using `update*` Functions  

  - Every input widget has a corresponding update function  
    + `updateActionButton()`  
    + `updateCheckboxInput()`  
    + `updateNumericInput()`  
    + ...
 
---

# Using `update*` Functions  

  - Every input widget has a corresponding update function  
    + `updateActionButton()`  
    + `updateCheckboxInput()`  
    + `updateNumericInput()`  
    + ...
    
  - Requires session argument on server() function  
```{r session,echo=TRUE,eval=FALSE}
server <- function(input, output, session) {
  ---#  do stuff
}
```  

- After all observers (reactive things) evaluate, updater sends message back to client



---

# Using `update*` Functions

- Syntax of `update*` functions similar to the functions that created the inputs

```{r update, echo = TRUE, eval = FALSE}
numericInput(inputId, 
             label, 
             value, 
             min = NA, 
             max = NA, 
             step = NA,
             width = NULL)

updateNumericInput(session, 
                   inputId, 
                   label = NULL, 
                   value = NULL,
                   min = NULL, 
                   max = NULL, 
                   step = NULL)  
```  



---

# Using `update*` Functions

- Syntax of `update*` functions similar to the functions that created the inputs
  
- Any arguments with `NULL` values ignored 

- For `radioButtons()`, `checkboxGroupInput()`, and `selectizeInput()`, the set of choices can be cleared by using `choices = character(0)` 

---

# `updateSliderInput()` (First Attempt)
```{r updateslider,eval=FALSE,echo=TRUE}
ui <- fluidPage(
      ...
      sidebarPanel(
         sliderInput("bins", "Number of bins:", 
                     min = 1, max = 50, value = 30),
         numericInput("maxBins", label = "Set Maximum Number of Bins",
                      value = 50, min = 1, max = 100)
      ),
      ...
)

server <- function(input, output, session) {
    ...
    updateSliderInput(session, "bins", max = input$maxBins)
}
```  
What is our issue?


---

# `updateSliderInput()` (Fixed)
```{r updateslider2,eval=FALSE,echo=TRUE}
ui <- fluidPage(
      ...
      sidebarPanel(
         sliderInput("bins", "Number of bins:",
                     min = 1, max = 50, value = 30),
         numericInput("maxBins", label = "Set Maximum Number of Bins",
                      value = 50, min = 1, max = 100)
      ),
      ...
)
server <- function(input, output, session) {
    ...
    observe({
      updateSliderInput(session, "bins", max = input$maxBins)
      })
}
``` 

---

# `update*` UI Functions  

 - Use the template app  
 
 - Try to add a numeric input for the user to specify the largest value of the slider  
 
 - Use the `updateSliderInput` function to update the max of the slider  
 
 - Add an `actionButton` to only update when pressed (so no intermediate updates while typing)  
  
 

---

# `renderUI()` and `uiOutput()`  

- `renderUI()` and `uiOutput()` allow for flexible dynamic UI elements

- Recall: Shiny essentially writes HTML/JavaScript for us!

```{r html,echo=TRUE,eval=TRUE}
print(fluidPage(titlePanel(title = "Hi"),
                sidebarLayout(sidebarPanel(), mainPanel())))
```


---

# `renderUI()` and `uiOutput()`  

- `renderUI()` and `uiOutput()` allow for flexible dynamic UI elements

- Recall: Shiny essentially writes HTML/JavaScript for us!

```{r html2,echo=TRUE,eval=TRUE}
print(numericInput("id", "Label User Sees", value = 10))
```

---

# `renderUI()` and `uiOutput()`  

- `renderUI()` makes a **reactive version** of a function that generates HTML!
      
- Can have `renderUI()` return

    + A widget or other function that makes HTML
    + A shiny 'tag object' created via `shiny::tagList()`
    
- Use with `uiOutput()` in UI file     

---

# `renderUI()` and `uiOutput()` (updating a widget)  

```{r renderui-widget,eval=FALSE,echo=TRUE}
  ui <- fluidPage(
        ...
        sidebarPanel(
            uiOutput("slider"),
            numericInput("maxBins", label = "Set Maximum Number of Bins",
                         value = 50, min = 1, max = 100)
        ),
        ...
  ),
  server <- function(input, output, session) {
      ...
      output$slider <- renderUI({
          sliderInput("bins", "Number of bins:", min = 1,
                      max = input$maxBins, value = 30)
      })
  }
``` 

---

# `renderUI()` and `uiOutput()` (outputting HTML)  

```{r renderui-infod,eval=FALSE,echo=TRUE}
  ui <- fluidPage(
        ...
        sidebarPanel(
             uiOutput("info"),
             numericInput("purchase", label = "How Many?",
                          value = 50, min = 0, max = 100)
        ),
        ...
  ),
  server <- function(input, output, session) {
      ...
        output$info <- renderUI({
          text <- paste0("You have selected to buy ", input$purchase)
          h3(text)
        })
  }
``` 

---

# `renderUI()` and `uiOutput()` (using `tagList()`)

```{r renderui-info,eval=FALSE,echo=TRUE}
  ui <- fluidPage(
        ...
        sidebarPanel(
             uiOutput("info"),
             numericInput("purchase", label = "How Many?",
                          value = 50, min = 0, max = 100)
        ),
        ...
  ),
  server <- function(input, output, session) {
      ...
        output$info <- renderUI({
          text <- paste0("You have selected to buy ", input$purchase)
          slidey <- sliderInput("bins", "Number of bins:", min = 1,
                      max = input$maxBins, value = 30)
          tagList(
            h3(text),
            slidey
            )
        })
  }
``` 


---

# `renderUI()` and `uiOutput()`   

- Use the template app  

- Try to add some dynamic updating text and things to the UI


---

# `conditionalPanel()`  

- Create a 'panel' that is only visible if a condition is met  

- Condition can depend on input widget
 
    + Accessed differently!  (Use a '.' not a '$' and javascript style comparisons)  

---

# `conditionalPanel()`  
```{r cond-panel-syntax,echo=TRUE,eval=FALSE}
...
sidebarPanel(
  selectInput("plotType", "Plot Type",
            c(Scatter = "scatter", Histogram = "hist")),
  
  # Only show this panel if the plot type is a histogram
  conditionalPanel(condition = "input.plotType == 'hist'",
          selectInput("breaks", "Breaks",
               c("Sturges", "Scott", "Freedman-Diaconis", "[Custom]" = "custom")),
      # Secondary conditonalPanel, Only show this panel if Custom is selected
      conditionalPanel(
          condition = "input.breaks == 'custom'",
          sliderInput("breakCount", "Break Count", min = 1, max = 200, value = 40)
      )
  )
)
```

---

# `conditionalPanel()`  

- Use the template app  

- Try to add a new UI element using a conditional panel!


---

# Dynamic UI Recap  
  
- Often want to update UI based on user input!  
  
- Methods for updating UI  
    + `update*` functions  
    + `renderUI()`/`uiOutput()`  
    + `conditionalPanel()`  

 
 
